---
title: "Data Science @ McDonald's"
subtitle: "Exercise 02 - Ultimate Big Mac Analysis"
author: "Khalil H Najafi"
date: "December 2019"
output: 
        ioslides_presentation:
                logo: mcd.png 
                widescreen: true
                smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(data.table)
library(reshape2)
library(plotly)

###### +--> Datasets ######

# Dataset containing forecasted units sold for selected items, weekly
# NB: for non ultimate Big Mac, forecast is based on non-promotional presence
dat_raw_forecast <- fread("./Data/data-forecast_units_sold_weekly.csv")

# Dataset containing actual units sold for selected items, weekly
dat_raw_actual <- fread("./Data/data-actual_units_sold_weekly.csv")

## Tidying
dat_forecast <- as_tibble(dat_raw_forecast) %>%
        melt(id.vars = "item") %>%
        rename(t = variable, units = value) %>%
        as_tibble()

dat_actual <- dat_raw_actual %>%
        melt(id.vars = "item") %>%
        rename(t = variable, units = value) %>%
        as_tibble()

# # # #

###### +--> Data Tables ######

## Non UBM Weekly Forecast Totals (non_ubm_wft) ; (f)
f <- dat_forecast %>%
        filter(item != "Ultimate Big Mac Limited Time Offer") %>%
        group_by(t) %>%
        summarise(total_units_fc = sum(units))

## Non UBM Weekly Actual Totals (non_ubm_wat) ; (g)
g <- dat_actual %>%
        filter(item != "Ultimate Big Mac Limited Time Offer") %>%
        group_by(t) %>%
        summarise(total_units_act = sum(units))

## Difference of Non UBM Weekly Actual to Forecast ; (δ)
d <- g %>%
        left_join(f) %>%
        mutate(d = total_units_act - total_units_fc) %>%
        select(t, d)

## Proportion of (absolute) difference to UBM actual ; (p)
p <- dat_actual %>%
        filter(item == "Ultimate Big Mac Limited Time Offer") %>%
        left_join(d) %>%
        mutate(p = abs(d)/units,
               p_inv = 1 - p) %>%
        select(t, p, p_inv)

## p and p_inv determine the success of the promotion based on a priori expectations/objectives
mean(p$p)
mean(p$p_inv)

## Comparing UBM actual v UBM forecast (dat_ubm)

#UBM forecasted
UBM_f <- dat_forecast %>%
        filter(item == "Ultimate Big Mac Limited Time Offer") %>%
        rename(units_f = units)

#UBM actual v UBM forecasted, by units and percentage
dat_UBM <- dat_actual %>%
        filter(item == "Ultimate Big Mac Limited Time Offer") %>%
        left_join(UBM_f) %>%
        mutate(delta = units - units_f,
               delta_perc = delta/units_f * 100)

#Average difference (percentage)
mean(dat_UBM$delta_perc)
```

## Context

This was a sample business case analysis featuring a limited time item, **The Ultimate Big Mac**

The purpose was to determine if the promotion was a (business) success

The datasets used are weekly sales data (in units) of the promotional item and a subset of other items, as well as forecasts of both the promotional item and notably the other items _in the absence of any promotion_; there are four weeks of data $(n = 4)$

A specific business objective was 35% of Ultimate Big Mac (UBM) units sold to be new/incremental, and what follows is the summary and break down of my analysis

## The Datasets

A brief review of the datasets: (1) actual units sold weekly; (2) forecasted units weekly 

```{r datasets, echo=F}
dat_raw_forecast

dat_raw_actual
```



## 1 - Did the UBM Meet the Business Objective?

In summary: yes, for the 4-week window the objective was exceeded

- The average proportion of UBM units attributable to new business $(p)$: **`r round(mean(p$p_inv)*100, 1)`%**
- The promotion exceeded the objective for each of the four weeks
    * Max was week 2 $(p^{max})$: **`r round(max(p$p_inv)*100, 1)`%**
    * Lowest was week 4 $(p^{min})$: **`r round(min(p$p_inv)*100, 1)`%**
    
## 1 - Meet Business Objective (cont) - Analysis Approach

To arrive at this, the process I followed was:

- Weekly actual without UBM and weekly forecast without UBM
    * $(weekly^{actual}_i - weeklyUBM^{actual}_i) = f_i$
    * $(weekly^{forecast}_i - weeklyUBM^{forecast}_i = g_i)$
- Take the difference of these to get the delta    
    * $f_i - g_i = \delta_i \longrightarrow \delta \epsilon (-\infty, +\infty)$
    * If delta is negative, then there is some loss/cannibalization from UBM, if delta is positive then UBM has improved general sales (multiplier .. though more likely forecasts have low confidence)
- Determine the proportion of delta to actual UBM units, and ensure it is less than 0.65
    * $\frac{|\delta_i|}{weeklyUBM^{actual}_i} = p_i \longrightarrow p_i \leq 0.65$
- Equivalently, check that one minus this proportion is greater than 0.35
    * $\therefore 1-p_i \geq 0.35$
    
## 1 - Meet Business Objective (cont) - Visualizations

Let's look at the weekly breakdown of proportion of UBM sales attributable to new business:

```{r Vis Weekly Prop, echo=F}
#mean p_inv
p_inv_bar <- mean(p$p_inv)
ggplotly(ggplot(p, aes(t, p_inv)) +
        geom_bar(stat = "identity",
                 fill = "#27251F") +
        geom_hline(yintercept = 0.35, linetype = "dashed", colour = "#DA291C") +
        annotate("text", x = 4.1, y = 0.37, label = "35% objective",
                 colour = "#DA291C", size = 3) +
        geom_hline(yintercept = p_inv_bar, colour = "#FFC72C") +
        annotate("text", x = 4.1, y = p_inv_bar+0.014, label = "Mean 51.2%",
                 colour = "#FFC72C", size = 3) +
        scale_y_continuous(labels = scales::percent) +
        labs(x = "",
             y = "1-p"))
```


## 1 - Meet Business Objective (cont) - Visualizations

Average proportion **~1.5X greater** than target: large variance in forecasted units to actual units for UBM

```{r Vis UBM Actual v Forecast, echo=F}
ggplotly(ggplot(dat_UBM, aes(t, units, group = item)) +
        geom_point(colour = "#DA291C") +
        geom_line(colour = "#DA291C") +
        geom_point(aes(t, units_f),
                   colour = "#FFC72C",
                   pch = 1) +
        geom_line(aes(t, units_f),
                  colour = "#FFC72C",
                  linetype = "dashed") +
        labs(x = "",
             y = "Units"))
```


## 2 - What Other Data Would Be Useful?

I would look at the following additions to the dataset initially:

- Full item units sold/forecasted
- Adding or identifying meals versus item-only purchases
- Date and location data (seasonality and local effects)
- Increased window $[n>4]$ (context and further seasonality/effects)
- Revenue data (ROI analysis)
    * this is useful as an input for partnering teams (Marketing, Research, etc) though should not fully dictate direction
    
The approach for the above would be to locate and aggregate the data into the datasets; expanding existing and/or additional tables


## 3 - Recommendations?

Business objective setting is no easy task, and fully considered takes more than revenue to inform, consider other dimensions:

- Experience
- Brand
- Competitive landscape (competitors offering certain combinations, "2 for $5")
- Food and eating trends (rise of breakfast items)

There may be generalizations from such promotions that are useful to abstract: a promotional sandwich, modification of a core/best selling item; this can be a useful input in designing campaigns or food ideas


## 4 - Expanding the Analysis

Aside from the additional variables mentioned previously, I was mostly curious about two somewhat related things:

- Comparing apples to apples (or Big Macs to Big Macs)
    * a more comparative basis is in the performance of the Ultimate Big Mac to other sandwich items such as the Big Mac, Quarter Pounder, McChicken, etc
    * if we cluster the menu list would there be categories that highlight more comparable items? How would these metrics change our conclusion? How would the objectives themselves change? (35% of new units may become 60% of average units within cluster)
- Exploring 'cannibalization': Are there relationships between items (positive/negative)?
    * it may be true that for some items, there are correlations with other items: are there more coffees added to McMuffin purchases? Are there no yogurts with Big Mac purchases? 
    * in determining this relationship, would there be a different result, for example, the availability of the UBM cannibalizes from McMuffin _and coffee_
    
## Thank You

Thoughts, feedback, questions .. did I take a completely wrong approach? :)



